---
title: "Vespa Velutina in Flanders"
author: "Federica Gazzelloni"
date: "2024-02-01"
page-navigation: true
format: 
  html:
    theme: cosmo
    code-overflow: wrap
    code-link: true
    code-copy: true
    code-fold: true
    code-summary: "Show the code"
    code-block-border-left: "#cdbdcc"
    code-block-bg: "gray90"
    fig-align: center
    fig-width: 8
    fig-height: 5
    linkcolor: "#88398a"
    link-external-newwindow: true   
    author-meta: "Federica Gazzelloni"
    toc: true
editor: visual
---

## GBif Data - download

Download `Vespa Velutina` data from [Global Biodiversity Information Facility (GBIF) for](https://www.gbif.org/) with the `{spocc}` package, which provide an R-interface to GBIF.

#### Load necessary libraries

```{r vespavelutina-1}
#| message: false
#| warning: false
# Load the libraries
library(tidyverse) # tools for data wrangling
library(spocc) # gbif R-interface for data download
library(giscoR) # Belgium polygons
library(ggspatial) # set the north arrow and the units bar
```


#### Download data from GBIF
```{r vespavelutina-2}
#| cache: true
#| code-fold: false
do_gbif <- spocc::occ("Vespa velutina",
                      from = "gbif",
                      date = c("2019-01-01","2021-01-01"),
                      limit = 10000,
                      has_coords = TRUE
                      )

do_gbif1 <- data.frame(do_gbif$gbif$data)
```



```{r vespavelutina-3}
#| echo: false
#| eval: false
do_gbif1%>%glimpse # Check the content of the download
```


```{r vespavelutina-4}
#| echo: false
# select the variable of interest and the final dimension
data <- do_gbif1%>%
  select(
    Vespa_velutina.name,
    Vespa_velutina.year,
    Vespa_velutina.longitude,
    Vespa_velutina.latitude,
    Vespa_velutina.coordinateUncertaintyInMeters,
    Vespa_velutina.continent,
    Vespa_velutina.country,
    Vespa_velutina.countryCode,
    Vespa_velutina.stateProvince
    ) 

# data %>% dim
```


```{r vespavelutina-5}
#| echo: false
# filter the data as needed and check the dimension
data1 <- data%>%
  filter(Vespa_velutina.continent=="EUROPE",
         Vespa_velutina.country=="Belgium",
         str_detect(Vespa_velutina.stateProvince,"Flanders"))%>%
  select(-Vespa_velutina.continent,
         -Vespa_velutina.country)

# data1%>%dim
```




#### Year-range of interest: 2019-2021
```{r vespavelutina-6}
data1%>%
  count(Vespa_velutina.year)
```

#### Observations by sub-location in Flanders: East and West Flanders 

```{r vespavelutina-7}
data1%>%count(Vespa_velutina.stateProvince)
```


#### Two sub-types of `Vespa velutina` are found in Flanders:

```{r vespavelutina-8}
data1%>%count(Vespa_velutina.name)
```



```{r vespavelutina-9}
#| echo: false
# set the theme options for the plots
theme_set(theme(
  text = element_text(size = 12, family = "Roboto Condensed"),
  plot.title = element_text(size=14,hjust = 0,face="bold"),
  plot.subtitle = element_text(hjust = 0),
  plot.caption = element_text(hjust = 0.5)
  ))
```



#### Download the Belgium regions polygons for mapping
```{r vespavelutina-10}
#| warning: false
#| message: false
#| cache: true
#| code-fold: false

# download Belgium regions polygons for mapping
be <- giscoR::gisco_get_nuts(resolution = "1", 
                     country = "BE", 
                     nuts_level = 2) 
```


## Map of Vespa velutina in Flanders

::: {.callout}
"Map tracking the distribution of Asian Hornets (Vespa velutina) in Flanders from 2019 to 2021, highlighting the spread of this invasive species and aiding conservation efforts."
:::

```{r vespavelutina-11}
#| warning: false
#| message: false
#| fig-align: "center"
ggplot()+
  # draw the be polygons
  geom_sf(data=be,fill="#e3e4bb")+ 
  # identify where the species is observed
  geom_point(data=data1, 
             aes(Vespa_velutina.longitude,Vespa_velutina.latitude,
                 fill=factor(Vespa_velutina.year)),
             shape=21,stroke=0.5,alpha=0.8,
             key_glyph=draw_key_rect,
             color="#bfc0ca")+
  # set a north arrow
  ggspatial::annotation_north_arrow(location="bl",
                                    pad_y = unit(1, "cm"),
                                    height = unit(0.8, "cm"),
                                    width = unit(0.8, "cm"))+
  # set a unit scale
  ggspatial::annotation_scale(pad_x = unit(0, "cm"),
                              pad_y = unit(0.5, "cm"))+
  # specify custom colors
  scale_fill_viridis_d(option = "H",
                       guide = guide_legend(
                         keyheight = unit(0.01, units = "mm"),
                         keywidth = unit(10, units = "mm")))+
  # set the title/subtitle and caption
  labs(title="Asian hornets (Vespa velutina) in Flanders (BE)",
       subtitle = "Occurrences by Year",fill="Year",
       caption = "\nDataSource: Global Biodiversity Information Facility (GBIF) | Map: Federica Gazzelloni\n")+
  # assure the proportion of the coordinates
  coord_sf(crs=4326)+
  # set a clean theme
  theme_void()
```

```{r vespavelutina-12}
#| echo: false
#| eval: false

# save the output as png
ggsave("vespavelutina_map.png",bg="#bfc0ca",
       height =5)
```

## Barplot of the observations by Year

```{r vespavelutina-13}
#| message: false
#| warning: false
#| fig-align: "center"

data1%>%
  ggplot(aes(factor(Vespa_velutina.year)))+
  geom_histogram(stat="count",fill="#e3e4bb")+
    labs(title="Asian hornets (Vespa velutina) in Flanders (BE)",
       subtitle = "Observations by Year",fill="Year",
       x="Time(Year)",
       caption = "\nDataSource: Global Biodiversity Information Facility (GBIF) | Map: Federica Gazzelloni\n")
```


```{r vespavelutina-14}
#| echo: false
#| eval: false
# save the output as png
ggsave("barplot.png",bg="#bfc0ca",height = 5)
```

